Making Sequential Programs Deterministic

### 2019-04-03 design.tex
Edits in section 2

@@

@@

\todo{Greekify the following}
\begin{itemize}
\item
Type polymorphism has the usual semantics.  For example, the type of the
identity function is $\forallt{\tau} \tau \rightarrow \tau$, which can be
equivalently written as
$\forallt{\kappa \ \beta} \kappa \ \beta \rightarrow \kappa \ \beta$.
(Compare this to 
$\forallt{\kappa} \forallt{\beta} \kappa \ \beta \rightarrow \kappa \ \beta$
which differs because the qualifier and basetype can vary independently;
see below.)\todo{Does the foregoing make sense?}

@@

  Basetype polymorphism is rarely used.  Full type polymorphism is more
  common, even if the type is decomposed and the parts used independently,
  as in this more general signature for the \<choose> function:

  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet> \beta$

@@

$\<length>\ :\ \forallt{\kappa} . \kappa\ \<String> \rightarrow \kappa \<Int>$

@@

$\<plus>\ :\ \forallt{\kappa} . \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa \<Int>$

@@

\OurTypeSystem uses the standard meaning of parametric
polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}.
A polymorphic abstraction (such as a class or function) is written and
typechecked once, but can be instantiated differently at different uses,
each of which behaves as if it was written with a different non-polymorphic
type.

Informally, the \<length> above acts as if it had multiple
(overloaded) definitions

@@

The body must typecheck at every possible instantiation, and at a use site,
the most specific applicable instantiation is chosen.
Here is an example:
(put it in a table)

@@

As described so far, polymorphism cannot express the collection behaviors
of \cref{sec:ond-behavior}.
Consider this typing for the \<size> function:

$\<size>\ :\ \forallt{\kappa} \kappa\ \<Collection> \rightarrow \kappa\ \<Int>$

It cannot be instantiated as

$\<size>\ :\ \<OrderNonDet>\ \<Collection> \rightarrow \<OrderNonDet> \<Int>$

because such an instantiation would include the invalid type \<OrderNonDet Int>.
A similar problem occurs for the \<first> operation that returns the first
element of a list.

\OurTypeSystem resolves this problem by introducing two variants to the
\<PolyDet> qualifier, \PolyDetUp\ and \PolyDetDown.
The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
other qualifiers unchanged.  The upward-pointing arrow is mnemonic of how
\<OrderNonDet> is changed:  it replaced by something higher in the type hierarchy.
The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
\<Det>, which is lower in the type hierarchy.
\Cref{fig:poly-resolutions} formalizes their behavior.

@@

$\<size> : \forallt{\kappa} \kappa\ \<Collection> \rightarrow \down{\kappa} \<Int>$

This can be instantiated at all three type qualifiers without creating any
invalid types, and the instantiations implement the semantics of \cref{sec:ond-behavior}:

$\<size> : \<NonDet>\ \<Collection> \rightarrow \<NonDet> \<Int>$

$\<size> : \<OrderNonDet>\ \<Collection> \rightarrow \<Det> \<Int>$

$\<size> : \<Det>\ \<Collection> \rightarrow \<Det> \<Int>$

An example use of $\uparrow$ is the \<first> operation.  A simplified
version of its type is

$\<first> : \forallt{\kappa\ \beta} \kappa\ \<List>\angles{\kappa\ \beta} \rightarrow \up{\kappa}\ \beta$

whose type qualifier can be instantiated as

$\<first> : \forallt{\beta} \<NonDet>\ \<List>\angles{\<NonDet>\ \beta} \rightarrow \<NonDet>\ \beta$

$\<first> : \forallt{\beta} \<OrderNonDet>\ \<List>\angles{\<OrderNonDet>\ \beta} \rightarrow \<NonDet>\ \beta$

$\<first> : \forallt{\beta} \<Det>\ \<List>\angles{\<Det>\ \beta} \rightarrow \<Det>\ \beta$

The actual type of \<first> accounts for more possible types for the type arguments.

@@

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution ($\uparrow$ is overloaded for types and qualifiers)}
\label{fig:poly-resolutions}
\end{figure}

@@

$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \<String>$

@@

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

$\<get> : \forallt{\kappa\ \beta} \<List>\angles{\kappa\ \beta} \times \kappa\ \<Int> \rightarrow \up{\kappa}\ \beta$

@@

If either the list or the index is non-deterministic, then the result is
nondeterministic:

@@

A similar typing does \emph{not} work for the \<set> function that sets an
element of a list, because it would permit a deterministic collection must
not be side-effected in non-deterministic ways.  Consider the following
type for

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
\TheDeterminismChecker performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes
 as shown in \cref{fig-sorting}.
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

    $\infer[\rulename{sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|Collections.sort(x)|}$

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule LIST SORT in \cref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

\subsection{Maps discussion}\label{maps}
A \<Map> is a set of mappings from keys to values. A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.

We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
\begin{itemize}
    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered according
    to the natural ordering of the keys. 
\end{itemize}

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

\begin{figure}
    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

@@

    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{\<Map> and \<Set> creation rules}
    \label{fig-creation-rules}
\end{figure}

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet Sets}. We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \|z = x.equals(y)| & \angles{ \kappa_{x} \ \beta_{x}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } & \angles{ \kappa_{y} \ \beta_{y}} \nsqsubseteq \angles{ \|OrderNonDet \ List| } }$

    $\infer[\rulename{map precision}]{\Gamma \vdash \|z : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta| & \|z = x.get(y)|}$

    \caption{Improving precision}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List} (An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

@@

\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
executions across machines. However, we make exceptions when the argument to this method is one of 
\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.

>>>>

This section first describes basic polymorphism over type qualifiers and
over basetypes (\cref{sec:basic-polymorphism}).  Then, it describes two
extensions.
One extension resolves \<OrderNonDet> to \<NonDet> or \<Det> where
needed (\cref{polymorphism-up-down}).
The other extension permits use of polymorphic (type) variables without
affecting binding of those variables --- that is, it affects which
instantiation of an abstraction is chosen at a use site 
(\cref{bindings-uses}).

@@

\subsubsection{Qualifier and basetype polymorphism}\label{sec:basic-polymorphism}

\OurTypeSystem supports parametric
polymorphism~\cite{Abadi:1989:FIM:77350.77373,Plotkin:1993:LPP:645891.671433}
over both classes and methods.
A polymorphic abstraction (a class or method) is written and
typechecked once.
It acts as if it has multiple different types, and each use site is
typechecked using the most specific applicable (instantiated,
non-polymorphic) type.

@@

\todo{Greekify the following}
\begin{itemize}
\item
Type polymorphism has the usual semantics.  For example, the type of the
identity function is $\forallt{\tau} \tau \rightarrow \tau$, which can be
equivalently written as
$\forallt{\kappa \ \beta} \kappa \ \beta \rightarrow \kappa \ \beta$.

@@

  Basetype polymorphism is rarely used on its own.  Full type polymorphism is more
  common, even if the type is decomposed and the parts used independently,
  as in this more general signature for the \<choose> function:

  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$

@@

$\<length>\ :\ \forallt{\kappa} \kappa\ \<String> \rightarrow \kappa\ \<Int>$

@@

$\<plus>\ :\ \forallt{\kappa} \kappa\ \<Int> \times \kappa\ \<Int> \rightarrow \kappa\ \<Int>$

@@

Informally, the \<length> method above acts as if it had multiple
(overloaded) definitions

@@

The body of \<length> must typecheck at every possible instantiation of its
polymorphic function type.  At a use site of \<length>,
the most specific applicable instantiation is chosen.
Here is an example:
\todo{(put it in a table)}

@@

As described so far, polymorphism cannot express the collection behaviors
of \cref{sec:ond-behavior}.
Consider this potential typing for the \<size> method in class \CollectionE:

$\<size>\ :\ \forallt{\kappa} \kappa\ \CollectionE \rightarrow \kappa\ \<Int>$

\noindent
It cannot be instantiated at $\kappa = \<OrderNonDet>$ as

$\<size>\ :\ \<OrderNonDet>\ \CollectionE \rightarrow \<OrderNonDet>\ \<Int>$

\noindent
because such an instantiation would include the invalid return type \<OrderNonDet Int>.
This means that the only two instantiations are at 
$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
size would be \<NonDet Int>.
%
A similar problem occurs for the \<first> operation that returns the first
element of a list.

\OurTypeSystem resolves this problem by introducing two operators over
polymorhic (type) variables, $\uparrow$ and $\downarrow$.
The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
other qualifiers unchanged.  The upward-pointing arrow is mnemonic for replacing
\<OrderNonDet> by something higher in the type hierarchy.
The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
\<Det>, which is lower in the type hierarchy.
\Cref{fig:poly-resolutions} formalizes their behavior.

@@

$\<size> : \forallt{\kappa} \kappa\ \CollectionE \rightarrow \down{\kappa}\ \<Int>$

\noindent
This can be instantiated at all three type qualifiers without creating any
invalid types:

$\<size> : \<NonDet>\ \CollectionE \rightarrow \<NonDet>\ \<Int>$

$\<size> : \<OrderNonDet>\ \CollectionE \rightarrow \<Det>\ \<Int>$

$\<size> : \<Det>\ \CollectionE \rightarrow \<Det>\ \<Int>$

\noindent
The above instantiations implement the semantics of \cref{sec:ond-behavior}.

An example use of $\uparrow$ is in the \<first> method in class \ListKB\ that
returns the first element of a list.  Its type is

$\<first> : \forallt{\kappa} \kappa\ \ListKB \rightarrow \up{\kappa}\ \beta_e$

\noindent
which can be instantiated as

$\<first> : \<NonDet>\ \ListKB \rightarrow \<NonDet>\ \beta_e$

$\<first> : \<OrderNonDet>\ \ListKB \rightarrow \<NonDet>\ \beta_e$

$\<first> : \<Det>\ \ListKB \rightarrow \<Det>\ \beta_e$

\todo{As I mentioned in email, $\<NonDet>\ \beta_e$ looks wrong to me,
  because $\beta_e$ may be a collection whose type arguments are not
  compatible with \<NonDet>.  Maybe just remark that the instantiation does
  not happen in that case.  If we try to adjust the type arguments to fit
  \<NonDet>, then unsoundness occurs.}

@@

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\|Det, NonDet|\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
\todo{I think this needs to do something to nested collections.  Or maybe
  just clarify that if there is a nested collection such that this would
  yield an invalid type, then the instantiation does not apply.  That might
  be cleanest.}
\label{fig:poly-resolutions}
\end{figure}

@@

$\<get> : \forallt{\kappa} \kappa\ \<List>\angles{\kappa\ \<String>} \times \kappa\ \<Int> \rightarrow \kappa\ \<String>$
\todo{Used to be $\up{\kappa}$ for return type, but I don't think that is needed.}
\todo{Oh, maybe it's because we need to be able to instantiate in the tye
  parameter differently?  How do we enable that?}

@@

$\<get> : \<NonDet List>\angles{\<NonDet String>} \times \<NonDet Int> \rightarrow \<NonDet String>$

$\<get> : \<Det List>\angles{\<Det String>} \times \<Det Int> \rightarrow \<Det String>$

@@

If either the list or the index is (order) non-deterministic, then the result is
nondeterministic:

@@

A similar typing does \emph{not} work for the \<set> function that sets an
element of a list, because it would permit a deterministic collection to
be side-effected in non-deterministic ways.  Consider the following
possible type for

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
\TheDeterminismChecker performs local type inference within method bodies.
It does not perform whole-program type inference.
This makes separate compilation possible.
It forces programmers to write specifications (type qualifiers) on methods,
which is good style and valuable documentation.

The local type inference is flow-sensitive.  That is, each variable may
have a different type on every line of the program, depending on
assignments to the variable.  The inferred type always respects any
declared type:  that is, it is always a refinement (a subtype) of the
declared type, if any.

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine, as shown in 
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible type argument for a \<Det> collection is \<Det>.

It is not permitted to

\TheDeterminismChecker has additional type refinement rules for the sorting
methods of \<List>, \<Collections>, and \<Arrays> classes
\begin{figure}
%     $\infer[\rulename{list sort}]{\Gamma \vdash x : \|Det \ List|\angles{\|Det| \ \beta}}{\Gamma \vdash x : \|OrderNonDet \ List|\angles{\|Det| \ \beta} & \|x.sort()|}$
%     
%     \bigskip

  $\infer[\rulename{sort}]
  {\Gamma \vdash x : \<Det List>\angles{\<Det>\ \beta}}
  {\Gamma \vdash x : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(x)>}$

@@

    \caption{Sorting type refinement rules.  \<sort> is a method that works
    by side effect and changes the value---and the type---of its receiver.}
\todo{These don't capture the notion of ``the type \emph{after} the sort call''.}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule \rulename{sort} in \cref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

@@

\subsection{Maps}\label{maps}
A map is a collection of keys--value pairs.
Like other collections, a map can be nondeterministic,
order-nondeterministic, or deterministic.
Analogously to set membership, lookups are deterministic in
order-nondeterministic and deterministic maps; in addition, the iteration
order is fixed in a deterministic map.

We explain why each of these is possible by considering three common
implementation strategies.

A hash table is order-nondeterministic, because hash codes are
nondeterministic in general.

A hash table can be augmented by a linked list recording insertion order,
so that iteration order is deterministic.

A map can be backed by a sorted collection, such as a tree.  Such a map is
either \<NonDet> (if \<NonDet> keys or values are inserted into it) or
\<Det>, but never \<OrderNonDet>.

\subsection{Determinism checker specific rules for precision}\label{precision}
We annotated the \codeid{equals()} method of the \<Set> class in the JDK as
\begin{verbatim}
PolyDet("down") boolean equals(PolyDet Set<E> this, PolyDet Object o)
\end{verbatim}
While this is sound, it is imprecise in the case of \codeid{OrderNonDet
  Set}s.\todo{Explain why.}
 We added the following rule shown in \cref{fig-precision-rules} to our checker to improve precision.
\begin{figure}
    \todo{add line break}
    $\infer[\rulename{set precision}]{\Gamma \vdash \|x.equals(y) : Det|}{\Gamma \vdash \|x| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{x} \ \beta_{x}}, \|y| : \|OrderNonDet| \ \|Set|\angles{ \kappa_{y} \ \beta_{y}} & \kappa_{x} \ \beta_{x} \nsqsubseteq \|OrderNonDet \ List| & \kappa_{y} \ \beta_{y} \nsqsubseteq \|OrderNonDet \ List| }$

@@

    $\infer[\rulename{map precision}]{\Gamma \vdash \|x.get(y) : Det|}{\Gamma \vdash \|x : OrderNonDet \ Map\angles{ Det \ \beta_k, \kappa_v\ \beta_v }, y : Det \ \beta|}$

    \caption{Improving precision}
\todo{In the \rulename{map precision} rule, should the \<Det> in the
  conclusion be $\beta_v$?}
    \label{fig-precision-rules}
\end{figure}
According to this rule, if the \<Set> receiver has type \codeid{@OrderNonDet Set<@Det List} and the argument to the \codeid{equals()} method is
also of the same type, the return type gets the qualifier \<Det>. Without this precision rule, the return type would have been instantiated as \<NonDet> which
is sound but imprecise. This rule is careful enough to not annotate the return type of \<equals()> as \<Det> if a type argument to either the
receiver or the argument happens to be an \codeid{OrderNonDet List}.
\todo{The following is the key!  Put in earlier.}(An order nondeterministic list does not provide any guarantee on the
equality as it is dependent on the iteration order).

@@

### 2019-04-03 java-design.tex
Edits in section 2

>>>>

\subsection{Dataflow analysis and type refinement}\label{sec:dataflow-java}

\todo{Relate this to defaulting.}

It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       // No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\cref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.

\subsection{Maps}\label{sec:maps-java}

 A \<Map> is deterministic if its \<entrySet> is deterministic.
In other words, iterating over this \<entrySet> is guaranteed to produce the same values in the exact same order across executions.
The \<Map> interface in Java doesn't define an \<iterator()> interface. Therefore it's sound to equate the determinism of the \<Set>
returned by \<entrySet()> to the determinism of the \<Map>. Interestingly, we found that every implementation of the \<Map> 
interface had the following property: if the \<Set> returned by \<entrySet()> has a determinism type $\kappa$, then
the \<Set>s returned by \<keySet()> and \<values()> also have the exact same determinism type $\kappa$.

We noticed the following properties with respect to determinism regarding the most widely used \<Map> implementations:
\begin{itemize}
    \item \<HashMap> can never guarantee fully deterministic iteration over its entries. A \codeid{Det HashMap} does not exist.
    \item \<LinkedHashMap> can have any of \<Det>, \<OrderNonDet>, or \<NonDet> types. Iterating over a \<LinkedHashMap> returns
    its entries in the order of their insertion. An \<OrderNonDet LinkedHashMap> can be created by calling its constructor with a \<HashMap>.
    \item \<TreeMap> can either be \<Det> or \<NonDet>. An \<OrderNonDet TreeMap> doesn't exist because the entries always ordered.
\end{itemize}

We prohibit the creation of a \codeid{Det HashMap} or an \codeid{OrderNonDet TreeMap} in our determinism checker.
\Cref{fig-creation-rules} formalizes these rules. Similar constraints apply to \<HashSet>s and \<TreeSet>s which are
also expressed by the last two rules of \cref{fig-creation-rules}.

\begin{figure}
    $\infer[\rulename{hashmap creation}]{\|new \ \kappa\ HashMap()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

    \bigskip

    $\infer[\rulename{treemap creation}]{\|new \ \kappa\ TreeMap()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$

    \bigskip

    $\infer[\rulename{hashset creation}]{\|new \ \kappa\ HashSet()|}{\kappa \in \{ \<OrderNonDet>, \<NonDet> \}}$

    \bigskip

    $\infer[\rulename{treeset creation}]{\|new \ \kappa\ TreeSet()|}{\kappa \in \{ \<Det>, \<NonDet> \}}$
    \caption{\<Map> and \<Set> creation rules}
    \todo{Are these type well-formedness rules?}
\todo{What about \<new TreeSet(myOrderedCollection<Det ...>)> which yields a
  deterministic set?}
    \label{fig-creation-rules}
\end{figure}

In the JDK, \<LinkedHashSet> is implemented as a subclass of \<HashMap>.
\todo{Talk about this design choice. Josh Bloch's book: "If the code surrounding the first declaration depended on LinkedHashSet's ordering policy,
    then it would be incorrect to substitute HashSet for LinkedHashSet in the declaration, because HashSet makes no guarantee concerning iteration order."}

\subsection{Precision improvements}\label{sec:precision-java}

\codeid{System.getProperty()} returns a \<NonDet> type since we can't provide any determinism guarantees for
executions across machines. However, we make exceptions when the argument to this method is one of 
\codeid{"line.separator", "path.separator", or "file.separator"} as they return OS dependent values which are guaranteed to be deterministic on that OS.
